Estimating Animal Abundance with N-Mixture Models Using the \pkg{R-INLA} Package for R

### 2017-04-24 meehan_et_al_inla_nmix_jss/article.tex
more tweaks

%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Timothy D. Meehan\\National\\Audubon Society \And
        Nicole L. Michel\\National\\Audubon Society \And
        Haavard Rue\\King Abdulla University\\of Science and Technology}
\title{Estimating Animal Abundance with N-Mixture Models Using the \pkg{R-INLA} Package for R}

@@

%% The address of (at least) one author should be given
\Address{
  Timothy D. Meehan\\
  National Audubon Society\\
  Boulder, Colorado USA\\
  E-mail: \email{tmeehan@audubon.org}\\
\\
  Nicole L. Michel\\
  National Audubon Society\\
  San Francisco, California USA\\
  E-mail: \email{nmichel@audubon.org}\\
  \\
  Haavard Rue\\
  King Abdulla University
  of Science and Technology\\
  Thuwal, Saudi Arabia\\
  E-mail: \email{haavard.rue@kaust.edu.sa }
}
%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In recent years, new sampling schemes and modeling approaches have enabled improved estimates of animal abundance that are less biased by non-detection \citep{Denes_Silveira_Beissinger_2015}. One such sampling scheme, termed a metapopulation design \citep{Kery_Royle_2010}, involves repeat visits in rapid succession to each of multiple study sites in a study area. If, during repeat visits, the population is assumed to be closed (no immigration, emigration, or reproduction; static abundance), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability. This detection probability can, in effect, be used to correct abundance estimates for imperfect detection \citep{Royle_2004}.

@@

\subsection[Analysis with JAGS]{Analysis with JAGS}
We first analyzed the simulated data using \proglang{JAGS}, via the \pkg{runjags} package. In defining the model, we specified a Negative Binomial distribution for the abundance component, and used vague normal priors for the intercepts and the global effects of the covariates of $\lambda$ and $p$.

@@

\subsection[Analysis with unmarked]{Analysis with unmarked}
Next, we prepared the simulated data for the \pkg{unmarked} analysis, which involved slight modification of the data in the sim.data object. Here, the count data was changed from a 3-dimensional $I*J*K$ array to a 2-dimensional $I*K$ row by $J$ column matrix.  Each static site level variable was duplicated and stacked $K$ times to form a single-column vector. A column vector was created to identify each year in the stacked data. The site-year variable, x.p.3.mean, was transformed from 2-dimensional matrix to a single column vector. Reformatted variables were then assembled in an \pkg{unmarked} data structure called an unmarked frame.

@@

\subsection[Analysis with R-INLA]{Analysis with R-INLA}
Next, we prepared data for \pkg{R-INLA}, which uses data formats similar to those used by \pkg{unmarked}. The object counts.and.count.covs is an \pkg{R-INLA} object that includes an $I*K$ row by $J$ column matrix of counts. Next comes a value of 1, which is turned into an $I*K$ length vector of ones to specify that the model has a global intercept for $\lambda$. Finally, there are two $I*K$ length vector of values for the two static site covariates of abundance, where the vector of values for $I$ sites is stacked $K$ times. In addition to the counts.and.count.covs object, we also defined x.p.1.inla, which is a copy of x.lam.1.inla, and x.p.3.inla, which is the $I*K$ length vector corresponding with x.p.3.mean.

@@

\subsection[Analysis with JAGS]{Analysis with JAGS}
Given the long processing time associated with \proglang{JAGS} models, we only ran 3000 iterations (no thinning, after 1000 adaptive and 1000 burn-in iterations) during each simulation. This number is not sufficient for drawing inference from marginal posteriors, but was sufficient for looking at qualitative patterns in posterior means. In each of the 50 iterations, the size of the effect for x.p.3 was drawn from a uniform distribution that ranged from -3 to 3. The results of the simulations are depicted in Fig. \ref{fig:fig2}.

@@

\subsection[Analysis with unmarked]{Analysis with unmarked}
The \pkg{unmarked} model was specified with linear effects of transect length (length), elevation (elev), and forest cover (forest) on abundance, and a linear effect and quadratic effect of sampling intensity (ivel) and sampling date (date), respectively, on detection. Both sampling intensity and sampling date were site-survey, or survey level covariates.

@@

\subsection[Analysis with R-INLA]{Analysis with R-INLA}
The \pkg{R-INLA} model used a similar model structure. The only difference was that the \pkg{R-INLA} model was not able to process site-survey covariates for detection, so averaged site level covariates were used, instead. These averages were attained using the rowMeans() function.

@@

%% acknowledgements
\section*{Acknowledgments}
We thank the following individuals for commenting on previous drafts of this manuscript.

@@

The key observation is that both the Poisson and the Binomial distribution can be evaluated recursively in $n$,
\begin{displaymath}
    \text{Poisson}(n; \lambda) = \text{Poisson}(n-1; \lambda) \frac{\lambda}{n}
\end{displaymath}
and
\begin{displaymath}
    \text{Binomial}(y; n, p) = \text{Binomial}(y; n-1, p) \frac{n}{n-y}(1-p),
\end{displaymath}
and then also for the Poisson-Binomial product
\begin{displaymath}
    \text{Poisson}(n ; \lambda) \; \text{Binomial}(y;  n, p)
    =
    \text{Poisson}(n-1; \lambda) \; \text{Binomial}(y; n-1, p)
    \frac{\lambda}{n-y}(1-p).
\end{displaymath}
If we define $f_i = \lambda(1-p)/i$ for $i=1, 2, \ldots$, we can make use of this recursive form to express the likelihood with a finite upper limit as
\begin{eqnarray}
    \text{Prob}(y) &=& \sum_{n = y}^{n_{\text{max}}}
                       \text{Poisson}(n ; \lambda)\;
                      \text{Binomial}(y;  n, p) \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{Binomial}(y; y, p)
                       \Big\{ 1 + f_1 + f_1f_2 +
                    \ldots
                      +f_1\cdots f_{n_\text{max}}
                       \Big\} \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{Binomial}(y; y, p)
                       \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}\nonumber
\end{eqnarray}
The log-liklihood can then be evaluated using the following simple \proglang{R}-code.
\begin{CodeInput}
R> fac <- 1; ff <- lambda * (1-p)
R> for(i in (n.max - y):1){
+    fac <- 1 + fac * ff / i
+  } 
R> log.L <- dpois(y, lambda, log = TRUE) + dbinom(y, y, p, log = TRUE) + 
+           log(fac)
\end{CodeInput}
Since this evaluation is recursive in decreasing $n$, we have to choose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is small. Note that we are computing \code{fac} starting with the smallest contributions, which are more numerically stable.

>>>>

%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Timothy D. Meehan\\National\\Audubon Society \And
        Nicole L. Michel\\National\\Audubon Society \And
        H{\aa}vard Rue\\King Abdulla University\\of Science and Technology}
\title{Estimating Animal Abundance with N-Mixture Models Using the \pkg{R-INLA} Package for R}

@@

%% The address of (at least) one author should be given
\Address{
  Timothy D. Meehan\\
  National Audubon Society\\
  Boulder, Colorado USA\\
  E-mail: \email{tmeehan@audubon.org}\\
\\
  Nicole L. Michel\\
  National Audubon Society\\
  San Francisco, California USA\\
  E-mail: \email{nmichel@audubon.org}\\
  \\
  H{\aa}vard Rue\\
  King Abdulla University
  of Science and Technology\\
  Thuwal, Saudi Arabia\\
  E-mail: \email{haavard.rue@kaust.edu.sa }
}
%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In recent years, new sampling schemes and modeling approaches have enabled improved estimates of animal abundance that are less biased by non-detection \citep{Denes_Silveira_Beissinger_2015}. One such sampling scheme, termed a metapopulation design \citep{Kery_Royle_2010}, involves repeat visits in rapid succession to each of multiple study sites in a study area. If, during repeat visits, the population is assumed to be closed (no immigration, emigration, or reproduction; i.e., static abundance), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability. This detection probability can, in effect, be used to correct abundance estimates for imperfect detection \citep{Royle_2004}.

@@

\subsection[Analysis with JAGS]{Analysis with \proglang{JAGS}}
We first analyzed the simulated data using \proglang{JAGS}, via the \pkg{runjags} package. In defining the model, we specified a Negative Binomial distribution for the abundance component, and used vague normal priors for the intercepts and the global effects of the covariates of $\lambda$ and $p$.

@@

\subsection[Analysis with unmarked]{Analysis with \pkg{unmarked}}
Next, we prepared the simulated data for the \pkg{unmarked} analysis, which involved slight modification of the data in the sim.data object. Here, the count data was changed from a 3-dimensional $I*J*K$ array to a 2-dimensional $I*K$ row by $J$ column matrix.  Each static site level variable was duplicated and stacked $K$ times to form a single-column vector. A column vector was created to identify each year in the stacked data. The site-year variable, x.p.3.mean, was transformed from 2-dimensional matrix to a single column vector. Reformatted variables were then assembled in an \pkg{unmarked} data structure called an unmarked frame.

@@

\subsection[Analysis with R-INLA]{Analysis with \pkg{R-INLA}}
Next, we prepared data for \pkg{R-INLA}, which uses data formats similar to those used by \pkg{unmarked}. The object counts.and.count.covs is an \pkg{R-INLA} object that includes an $I*K$ row by $J$ column matrix of counts. Next comes a value of 1, which is turned into an $I*K$ length vector of ones to specify that the model has a global intercept for $\lambda$. Finally, there are two $I*K$ length vector of values for the two static site covariates of abundance, where the vector of values for $I$ sites is stacked $K$ times. In addition to the counts.and.count.covs object, we also defined x.p.1.inla, which is a copy of x.lam.1.inla, and x.p.3.inla, which is the $I*K$ length vector corresponding with x.p.3.mean.

@@

\subsection[Analysis with JAGS]{Analysis with \proglang{JAGS}}
Given the long processing time associated with \proglang{JAGS} models, we only ran 3000 iterations (no thinning, after 1000 adaptive and 1000 burn-in iterations) during each simulation. This number is not sufficient for drawing inference from marginal posteriors, but was sufficient for looking at qualitative patterns in posterior means. In each of the 50 iterations, the size of the effect for x.p.3 was drawn from a uniform distribution that ranged from -3 to 3. The results of the simulations are depicted in Fig. \ref{fig:fig2}.

@@

\subsection[Analysis with unmarked]{Analysis with \pkg{unmarked}}
The \pkg{unmarked} model was specified with linear effects of transect length (length), elevation (elev), and forest cover (forest) on abundance, and a linear effect and quadratic effect of sampling intensity (ivel) and sampling date (date), respectively, on detection. Both sampling intensity and sampling date were site-survey, or survey level covariates.

@@

\subsection[Analysis with R-INLA]{Analysis with \pkg{R-INLA}}
The \pkg{R-INLA} model used a similar model structure. The only difference was that the \pkg{R-INLA} model was not able to process site-survey covariates for detection, so averaged site level covariates were used, instead. These averages were attained using the rowMeans() function.

@@

%% acknowledgements
\section*{Acknowledgments}
We thank Jerry Cole and Curtis Burkhalter for commenting on previous drafts of this manuscript.

@@

The key observation is that both the Poisson and the Binomial distribution can be evaluated recursively in $n$,
\begin{displaymath}
    \text{Poisson}(n; \lambda) = \text{Poisson}(n-1; \lambda) \frac{\lambda}{n}
\end{displaymath}
and
\begin{displaymath}
    \text{Binomial}(y; n, p) = \text{Binomial}(y; n-1, p) \frac{n}{n-y}(1-p),
\end{displaymath}
and then also for the Poisson-Binomial product
\begin{displaymath}
    \text{Poisson}(n ; \lambda) \; \text{Binomial}(y;  n, p)
    =
    \text{Poisson}(n-1; \lambda) \; \text{Binomial}(y; n-1, p)
    \frac{\lambda}{n-y}(1-p).
\end{displaymath}
If we define $f_i = \lambda(1-p)/i$ for $i=1, 2, \ldots$, we can make use of this recursive form to express the likelihood with a finite upper limit as
\begin{eqnarray}
    \text{Probability}(y) &=& \sum_{n = y}^{n_{\text{max}}}
                       \text{Poisson}(n ; \lambda)\;
                      \text{Binomial}(y;  n, p) \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{Binomial}(y; y, p)
                       \Big\{ 1 + f_1 + f_1f_2 +
                    \ldots
                      +f_1\cdots f_{n_\text{max}}
                       \Big\} \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{Binomial}(y; y, p)
                       \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}\nonumber
\end{eqnarray}
The log-liklihood can then be evaluated using the following simple \proglang{R}-code.
\begin{CodeInput}
R> fac <- 1; ff <- lambda * (1-p)
R> for(i in (n.max - y):1){
+    fac <- 1 + fac * ff / i
+  } 
R> log.L <- dpois(y, lambda, log = TRUE) + dbinom(y, y, p, log = TRUE) + 
+           log(fac)
\end{CodeInput}
Since this evaluation is recursive in decreasing $n$, we have to choose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is small. Note that we are computing \code{fac} starting with the smallest contributions, which are more numerically stable.
